---
title: "Decreased Alcohol Consumption in Britain"
author: "Ahmed Tadde"
date: "June 21, 2017"
output:
   html_document:
      self_contained: false
      keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
A survey conducted by the Office for National Statistics in Britain reveals adult drinking habits. The collected data contains information on the proporation of respondents who drank in the last week, or drank on 5 or more days in the last week, or practice [teetolism](https://en.wikipedia.org/wiki/Teetotalism>). The (simplfied version?) data can be downloaded [here](https://onedrive.live.com/view.aspx?resid=43EBDBC5D5265516!11641&ithint=file%2cxlsx&app=Excel&authkey=!AIN6rnJ4jqWf7Kk).
The following sections consist of some light data cleaning/standarization and few visualizations on the trends of drinking habits among people of different ages and gender.

## Libraries & Functions

### Libraries
```{r}
##=====  Loading Libraries for Processing
# load pacman first; then from pacman, use p_load to load other libraries
# This is so that p_load can automatically install the package if it's not
# already installed => efficient fulfillment of package requirements!

library(pacman)
p_load(data.table)
p_load(foreach)
p_load(plotly)
p_load(dplyr)
p_load(stringi)
```

### Functions
```{r}
##===== Functions for Data Wrangling

ReformatYear <- function(values){
  #takes in vector of Year values (char type)
  #removes all occurences of ","
  #returns modified vector of Year values (char type)

  foreach(i=1:length(values), .combine = c) %do%{
    values[i] <- as.character(gsub(",","", values[i]))
  }
}


ReformatGender <- function(values){
  #takes in vector of Gender values (char type)
  #changes all occurences of "All persons" => 'All'
  #returns modified vector of Gender values (char type)
  
  values <- stri_trim_both(as.character(values))
  old <- 'All persons'
  new <- 'All'
  values[which(values %in% old)] <- new
  values
}


ReformatAgeRange <- function(values){
  #takes in vector of AgeRange  values (char type)
  #changes all occurences of "All 16+" => '16+'
  #returns modified vector of AgeRange values (char type)
  
  values <- stri_trim_both(as.character(values))
  old <- 'All 16+'
  new <- '16+'
  values[which(values %in% old)] <- new
  values
}


ReformatProportion <- function(values){
  #takes in vector of Proportion values (numeric type)
  #removes all occurences of "%"
  #returns modified vector of Proportion values (numeric type)

  foreach(i=1:length(values), .combine = c) %do%{
    values[i] <- as.numeric(gsub("%","", values[i]))
  }
}


ReformatQuestion <- function(values){
  #takes in vector of Question values (char type)
  #Changes: 
  #'Drank alcohol in the last week' => 'Drank alcohol'
  #'Drank alcohol on five on more days in the last week' => ''Drank alcohol on 5+ days'
  #returns modified vector of Question values (char type)
 
  old <- 'Drank alcohol on five on more days in the last week'
  new <- 'Drank alcohol on 5+ days'
  values[which(values %in% old)] <- new
  
  old <- 'Drank alcohol in the last week'
  new <- 'Drank alcohol'
  values[which(values %in% old)] <- new
  
  values
}



MakeBarPlot <- function(df){
  #takes in a dataframe, all columns included
  #returns bar plots: Question vs Porportion BY 2005,2016
  
 df[
    Year %in% c(2005,2016) & AgeRange == '16+' & Gender == 'All',
    .(Year, Question, Proportion)
  ][, 
    Change := c(NA,diff(Proportion)), by= 'Question'
  ] -> table
  
  use.font <- list(
    family = "Courier New, monospace",
    size = 15,
    color = "black"
  )
  
  ax <- list(
    title = "",
    showticklabels = TRUE,
    zeroline = FALSE,
    showline = FALSE,
    showgrid = FALSE,
    tickfont = use.font
  )
  
  ay <- list(
    title = '',
    titlefont = use.font,
    showticklabels = FALSE,
    zeroline = FALSE,
    showline = FALSE,
    showgrid = FALSE
  )
  
  plot_ly(
    type = 'bar',
    table,
    x = ~Question,
    y = ~Proportion,
    color = ~Year, colors = c("#66757F","#32AB9F"),
    hoverinfo = 'text',
    text = ~paste0(
      "<br>",'Proportion: ', Proportion,'%'
    )
  )%>%layout(
    title="Alcohol Consumption Drops in Britain", titlefont = use.font, xaxis=ax, yaxis=ay ,
    annotations = list(
      x = ~Question,
      y = ~Proportion+10,
      text = c('', '-7.3%', '', '-7.2%', '', '+2%'),
      xanchor = 'center',
      yanchor = 'bottom',
      showarrow = FALSE
    ),
    legend = list(x = 0.6, y = 0.7, orientation = 'h')
  )
  
  
  
}

MakeLinePlot <- function(df, age.range, question, counter){
  #takes in a dataframe,
  #returns a series plot: ProportionChange vs Year
  
  table <- df[AgeRange %in% age.range & Question %in% question]
  
  use.font <- list(
    family = "Courier New, monospace",
    size = 12,
    color = "black"
  )
  
  ax <- list(
    title = "",
    showticklabels = TRUE,
    zeroline = FALSE,
    showline = FALSE,
    showgrid = FALSE,
    tickfont = use.font
  )
  
  ay <- list(
    title = question,
    titlefont = use.font,
    showticklabels = FALSE,
    zeroline = FALSE,
    showline = FALSE,
    showgrid = FALSE
  )
  
  if (counter != 3){
    
    table %>% plot_ly(
      height = 600,
      width = 800,
      x = ~Year,
      y = ~ProportionChange,
      type = "scatter",
      mode = "lines",
      color = ~Gender,
      colors = c("#008080","#708090","#B8860B"),
      hoverinfo ="text",
      showlegend = FALSE,
      text =~paste0(
        '<br>', question,
        '<br>', 'Gender: ', Gender,
        '<br>', 'Change: ', ProportionChange,"%",
        '<br>', 'Year: ', Year
        
      )
    )%>%
      layout(
        title = age.range, 
        titlefont = use.font , 
        xaxis = ax, 
        yaxis = ay,
        autosize = F
      )
  }else{
    table %>% plot_ly(
      height = 600,
      width = 800,
      x = ~Year,
      y = ~ProportionChange,
      type = "scatter",
      mode = "lines",
      color = ~Gender,
      colors = c("#008080","#708090","#B8860B"),
      hoverinfo ="text",
      # showlegend = TRUE
      text =~paste0(
        '<br>', question,
        '<br>', 'Gender: ', Gender,
        '<br>', 'Change: ', ProportionChange,"%",
        '<br>', 'Year: ', Year
        
      )
    )%>%
      layout(
        title = paste0('AGE RANGE: ', age.range), 
        titlefont = use.font , 
        xaxis = ax, 
        yaxis = ay,
        autosize = F
      )
  }
  
  
  
}

FetchLinePlots <- function(df){
  #takes in a dataframe
  #returns a nested list of plots... by Age Range, then by Question
  
  age.ranges <- unique(df$AgeRange)
  questions <- unique(df$Question)
  
  foreach(i=1:length(age.ranges)) %do%{
    return(
      foreach(j=1:length(questions)) %do%{
        return(MakeLinePlot(df, age.ranges[i], questions[j], j))
      }
    )
  } -> all.plots
  
  names(all.plots) <- age.ranges
  
  
  return(all.plots)
}
```



## Data Cleaning

### Loading & Glimpse
```{r}
data <- fread('data.csv')

glimpse(data)
unique(data$Year)
table(data$`Age Range`)
table(data$Gender)
table(data$Question)


```


### Cleaning 
```{r, echo=TRUE, results= 'hide'}
####===== Reformating columns
data[, Year:= ReformatYear(Year)]
data[, Gender := ReformatGender(Gender)]
data[, Proportion := ReformatProportion(Proportion)]
data[, Question := ReformatQuestion(Question)]

####===== Change Age Range => AgeRange AND REFORMAT
setnames(data, "Age Range", "AgeRange")
data[, AgeRange := ReformatAgeRange(AgeRange)]

####===== Create PoportionChange variable
data[, ProportionChange := round(c(0,diff(Proportion)),2), by = c("Question", "AgeRange","Gender")]

```



## Visualizations

### Overall Trend

According to data, compared to 2005, Britons are drinking less in 2016. In contrast, the number of teetotalers has increased.
```{r}

####===== Bar Plots By Question showing proportion difference for 2005 & 2016
MakeBarPlot(data)


```





### Changes by Categories

The trend lines below show the changes in the proportion of people consuming alcohol since 2005
```{r}

####===== Make Line Plots for Proportion Changes
line.plots <- FetchLinePlots(data)

```



```{r, echo=FALSE}
plotly::subplot(
  line.plots$`16 to 24`,
  nrows = 3,
  shareX = TRUE, shareY = TRUE
)

```


```{r, echo=FALSE}
plotly::subplot(
  line.plots$`25 to 44`,
  nrows = 3,
  shareX = TRUE, shareY = TRUE
)%>% layout(showlegend =  FALSE)

```


```{r, echo=FALSE}
plotly::subplot(
  line.plots$`45 to 64`,
  nrows = 3,
  shareX = TRUE, shareY = TRUE
)%>% layout(showlegend =  FALSE)

```


```{r, echo=FALSE}
plotly::subplot(
  line.plots$`65+`,
  nrows = 3,
  shareX = TRUE, shareY = TRUE
) %>% layout(showlegend =  FALSE)

```



```{r, echo=FALSE}
plotly::subplot(
  line.plots$`16+`,
  nrows = 3,
  shareX = TRUE, shareY = TRUE
)%>% layout(showlegend =  FALSE)

```

# Final Thoughts
What are possible reasons for the observed decreased in alcohol consumption? Are Britons becoming more
health contious? If that is the case, what other habits are on the decline? Would data from Germany and accross Europe reveal an overall trend of decreased alcohol consumption? 



